---
title: "STRIDE: Simulating Travel Rates In Diverse Environments"
author: "Mickey Campbell"
date: last-modified
format: gfm
toc: true
---

## Introduction

`stride` is an R library for mapping least-cost paths and estimating travel time across a diverse range of landscape conditions. It employs the Simulating Travel Rates In Diverse Environments (STRIDE) model, which was recently developed as part of a study by Campbell et al. (*in review*):

Campbell, Michael J., Cutler, Sierra L., Dennison, Philip E., (*in review*). A singular, broadly-applicable model for estimating on- and off-path walking travel rates using airborne lidar data.

The study employed the use of three different field experiments, where three different groups of study subjects walked a series of 100 m-long linear transects through wide-ranging environmental conditions, from flat urban streets to densely vegetated, steep forests with unimproved ground surfaces. The travel rates from these experiments were compared to lidar-derived estimates of terrain slope, low-lying vegetation density, and ground surface roughness within each of the transects using non-linear least-squares regression. The resulting model demonstrated an impressive ability to accurately predict travel rates across wide-ranging landscape conditions, and yielded logically consistent least-cost paths when applied in a spatial context.

To enable broad use and consistent implementation of STRIDE for travel time estimation and least-cost path mapping, we developed `stride`. This open-source R library allows users to apply STRIDE in a simple and user-friendly way. All the user needs is a lidar point cloud dataset and a set of some origin and destination points, and `stride` can map the least-cost paths between them and calculate an accurate travel time estimate.

## Installing `stride`

The first step in the STRIDE modeling workflow is to install the `stride` R package. At present, `stride` is not available on CRAN, so you will need to install from Github. One way to do this is shown below:

```{r}
#| eval: false
#| warning: false
#| message: false

# install using devtools
devtools::install_github("https://github.com/mickeycampbell/stride")
```

## Loading necessary libraries

In addition to `stride`, we will employ the use of several other useful R libraries to ensure successful execution of the STRIDE modeling workflow. If you do not have these libraries installed already, you will need to do so prior to loading them. Since they are all available on CRAN, you can simply use `install.packages()`.

```{r}
#| warning: false
#| message: false

library(stride)
library(lidR)
library(terra)
```

## Downloading lidar data

To test the STRIDE modeling workflow, you will need an airborne lidar dataset to work with. There are many potential sources of free airborne lidar data in the US. For the sake of this example, we will use data from the [United States Geological Survey (USGS) 3D Elevation Program (3DEP)](https://www.usgs.gov/3d-elevation-program). One approach for exploring 3DEP data availability and downloading is through [The National Map](https://apps.nationalmap.gov/downloader/#/). To see where lidar data are available, you can check the box next to *Elevation Source Data (3DEP) - Lidar, IfSAR*. Under *Subcategories*, you can check the box next to *Lidar Point Cloud (LCP)*. By clicking *Show*, a layer will be added to the map interface displaying where lidar data are available for download, colored by [Quality Level](https://www.usgs.gov/3d-elevation-program/topographic-data-quality-levels-qls).

To download data within an area of interest, you have several options, including drawing an extent (bounding box), polygon, or point, entering in extent coordinates manually, or uploading a shapefile or KML. In this example, we will download four adjacent tiles of lidar data in a mountainous, forested area near Brighton, UT. 3DEP lidar data are typically delivered in 1x 1km tiles in LAZ format -- a compressed point cloud data type. So, the example study area will be approximately 2x2 km in size.

## Generating a DTM

Once you have your lidar data downloaded, the first major step in the STRIDE workflow is generate a digital terrain model (DTM). To do this, you will need to use the `lidR` library. `lidR` is an awesome R library that incorporates a vast array of lidar processing, analysis, and visualization functions. One of the great strengths of `lidR` is its ability to process data in parallel. Because lidar data are so massive, data are almost always tiled into chunks. `lidR`'s `LAScatalog` object type allows you to read and process tiled lidar data in parallel, using as many cores as your computer has (or as many as your RAM will alow). Since we have four tiles of lidar data, we can process those tiles in parallel by reading the data in as a `LAScatalog` using `readLAScatalog()`. If, instead, you preferred to read in a single tile of data, you could use `readLAS()`.

Once the data are read in, we will use `gen_dtm()` to generate DTMs at two different spatial resolutions:

1. **1 m resolution**: This DTM will be used for generating a ground surface roughness raster dataset, which requires modeling the terrain at a high spatial resolution.
2. **10 m resolution**: This DTM will be used as the basis of calculating directional slope, which is one of the movement costs that are factored into STRIDE.

```{r}
#| warning: false
#| message: false

# suppress progress bars
options(lidR.progress = F)

# read in the lidar data as a lascatalog
ctg <- readLAScatalog("C:/temp/stride/lidar_data/raw",
                      filter = "-drop_withheld -drop_class 7 18")
ctg
plot(ctg)
```

As you can see, we filtered out points classified as being "withheld" or belonging to one of the two "noise" classes (7 and 18) -- this aims to remove bad data from the analysis. By calling `ctg` after reading the data in, we can see some basic information about the four tiles of lidar data in this example. By calling `plot(ctg)`, we can see the spatial extents of the four tiles.

Next, we will generate our DTMs.

```{r}
#| warning: false
#| message: false

# generate 1m DTM
dtm_1m <- gen_dtm(
  las = ctg,
  res = 1,
  ncores = 4L,
  out_file = "C:/temp/stride/lidar_data/dtm_1m.tif",
  overwrite = T
)
dtm_1m

# generate 10m DTM
dtm_10m <- gen_dtm(
  las = ctg,
  res = 10,
  ncores = 4L,
  out_file = "C:/temp/stride/lidar_data/dtm_10m.tif",
  overwrite = T
)
dtm_10m

# plot out the 10m DTM
plot(dtm_10m)
```

## Generating a ground surface roughness raster

One of the three cost surfaces that STRIDE relies on to estimate travel impedance/conductance and enable the mapping of least-cost paths and accurate estimation of travel times is the roughness of the ground surface. In the context of STRIDE, roughness is calculated as the mean of absolute differences between a fine-scale (1 m) DTM and a "smoothed" version of that DTM, where the smoothing is the result of a circular mean focal filter. By default, the filter has a radius of 2 m. To generate this raster, we will use the `gen_rgh()` function.

```{r}
#| warning: false
#| message: false

# generate a ground surface roughness raster
rgh <- gen_rgh(
  dtm = dtm_1m,
  radius = 2,
  res = 10,
  out_file = "C:/temp/stride/lidar_data/rgh.tif",
  overwrite = T
)
rgh

# plot it out
plot(rgh)
```

## Generating a vegetation density raster

Another of the three cost surfaces that STRIDE relies on is the density of vegetation within the range of human height. In Campbell et al. (*in review*), the height range that was most highly correlated with travel impedance was 0.85 - 1.20 m. For an average individual 1.70 m tall, this is approximately waist to chest height -- vegetation in this range is difficult to avoid, and thus a high density of it can slow you down considerably. Lidar data are excellent at quantifying and mapping vegetation structure. Because they are measured using narrow laser pulses that can exploit small gaps in a forest canopy, even in the presence of tall vegetation, the density of vegetation near the ground surface can be readily quantified. To capture vegetation density, we rely on a lidar poin cloud-derived metric referred to as the **normalized relative density (NRD)**:

$$NRD=\frac{\sum_{i}^{j}n}{\sum_{0}^{j}n}$$

where $\sum_{i}^{j}n$ is the number ($n$) of lidar points between heights $i$ and $j$ and $\sum_{0}^{j}n$ is the number ($n$) of lidar points between 0 m (the ground level) and $j$. It is a relative measure of point density that accounts for canopy occlusion that measures the proportion of energy intercepted by the vegetation within a height range of interest prior to reaching the ground.

To calculate NRD, however, requires that lidar points are **normalized to the ground surface**. By default, of course, lidar data *z* values are measured in elevation above mean sea level. Accordingly, we need to convert elevations to heights above the ground. This can be done using the `norm_hgt()` function.

```{r}
# normalize lidar point heights
las_hgt <- norm_hgt(
  las = ctg,
  ncores = 4L,
  out_path = "C:/temp/stride/lidar_data/hgt"
)
las_hgt
```

To make sure this worked, we can take a look at a random sample of *z* values from one of our height-normalized point clouds:

```{r}
# read in one of the height-normalized lidar tiles
laz_files <- list.files("C:/temp/stride/lidar_data/hgt", full.names = T)
test_file <- laz_files[1]
test_las <- readLAS(test_file)
test_zs <- sample(test_las$Z, 1000)
hist(x = test_zs, xlab = "Height (m)", ylab = "Frequency")
```

Looks about right! It's not uncommon to have some slightly negative values -- height normalization is an imperfect process. This will not affect the analysis in any significant way.

With our height-normalized data, we can now calculate NRD. To do this, we can use the `gen_dns()` function. We will use the default values from Campbell et al. (*in review*) to capture the optimal height range from that study (0.85 - 1.20 m) and recommend that you do the same.

```{r}
# generate a vegetation density raster
dns <- gen_dns(
  las_hgt = las_hgt,
  hgt_1 = 0.85,
  hgt_2 = 1.20,
  res = 10,
  ncores = 4L,
  out_file = "C:/temp/stride/lidar_data/dns.tif",
  overwrite = T
)
dns

# plot it out
plot(dns)
```

## Checking raster alignment

An important step in the STRIDE modeling workflow is ensuring that all of your cost rasters are spatially congruent. That is, they share the same coordinate reference system (CRS), spatial resolution, extent, and origin. This can be done easily with `algn_rasts()`. This function first checks to see if these spatial characteristics are congruent. Optionally, it can also force them into congruence. This option should be employed cautiously, as the user has little control over how that congruence is forced. It may be beneficial to manually force congruence using tools in the `terra` library. Provided that the entire workflow has taken place as outlined above, the three inputs should be spatially congruent and not require forcing. Only if rasters were generated outside of this package might this become necessary.

```{r}
# check raster alignment
r <- algn_rasts(
  dtm = dtm_10m,
  dns = dns,
  rgh = rgh,
  force = T
)
```

## Get barrier rasters

In addition to the three cost rasters (DTM, vegetation density, and ground surface roughness), STRIDE was formulated with some barriers in mind, to capture landscape features that inhibit pedestrian movement. The function `get_bars()` creates rasterized representation of two types of barriers:

1. **Cliffs**: Areas with terrain slopes greater than 45 degrees.
2. **Waterbodies**: Ponds and lakes that one cannot walk across.

Cliffs are generated directly from the 10 m DTM. Waterbodies can come from one of two sources: (1) user-supplied; or (2) automatically downloaded from the [USGS National Hydrography Dataset (NHD) Plus](https://www.usgs.gov/national-hydrography/nhdplus-high-resolution). The NHD is only available in the US, so users outside of the US will need to supply their own hydrography data. Users within the US can also supply their own hydrography data, but if they do not, waterbodies will be downloaded from the USGS NHD. Of course, this means an internet connection is required for successful use.

The barrier rasters will automatically be aligned with the cost rasters, so no additional congruency checks are needed.

```{r}
# get barrier raster data
bars <- get_bars(
  dtm = dtm_10m,
  out_file_clf = "C:/temp/stride/lidar_data/clf.tif",
  out_file_wtr = "C:/temp/stride/lidar_data/wtr.tif",
  overwrite = T
)
clf <- bars$clf
clf
wtr <- bars$wtr
wtr

# plot them out
plot(clf)
plot(wtr)
```

Cells with a value of 0 are considered barriers in each of these datasets, since transition matrices rely on the concept of **conductance** -- the opposite of impedance. Therefore, areas with 0 conductance represent impassible features.

## Build transition matrix

The STRIDE modeling workflow leans heavily on the `gdistance` package -- an excellent R package for modeling least-cost paths in an efficient and flexible way. `gdistance` relies on the creation of **transition matrices**, which are sparse matrix representations of conductance between neighboring cells in a raster dataset. The function `bld_tm()` builds a STRIDE transition matrix using the three cost rasters and two barrier rasters.

```{r}
# build a transition matrix
tm <- bld_tm(
  dtm = dtm_10m,
  dns = dns,
  rgh = rgh,
  clf = clf,
  wtr = wtr
)
tm
```

## Map least-cost paths

With the transition matrix now built, we can now map the least cost path(s) between two (or more) points within the study area. In `stride`, this can be accomplished using the `map_lcp()` function. In the example below, we will simply generate two random points and map the least-cost path in both directions between them.

```{r}
# set random seed
set.seed(1234)

# generate random start and end points
pt_orig <- spatSample(
  x = dtm_10m,
  size = 1,
  as.points = T,
  values = F
)
pt_dest <- spatSample(
  x = dtm_10m,
  size = 1,
  as.points = T,
  values = F
)

# map least-cost path and compute travel time
lcp_1 <- map_lcp(
  tm = tm,
  pt_orig = pt_orig,
  pt_dest = pt_dest
)
lcp_2 <- map_lcp(
  tm = tm,
  pt_orig = pt_dest,
  pt_dest = pt_orig
)

# fix missing crs
crs(lcp_1) <- crs(dtm_10m)
crs(lcp_2) <- crs(dtm_10m)

# plot out the data
plot(dtm_10m, main = "Elevation")
plot(lcp_1, add = T, col = "red")
plot(lcp_2, add = T, col = "blue")
plot(pt_orig, add = T)
plot(pt_dest, add = T)

plot(dns, main = "Density")
plot(lcp_1, add = T, col = "red")
plot(lcp_2, add = T, col = "blue")
plot(pt_orig, add = T)
plot(pt_dest, add = T)

plot(rgh, main = "Roughness")
plot(lcp_1, add = T, col = "red")
plot(lcp_2, add = T, col = "blue")
plot(pt_orig, add = T)
plot(pt_dest, add = T)

# get total time
print(paste0("Total travel time for LCP 1: ", round(lcp_1$time), " sec"))
print(paste0("Total travel time for LCP 2: ", round(lcp_2$time), " sec"))

# get total length
print(paste0("Total travel time for LCP 1: ", round(perim(lcp_1)), " m"))
print(paste0("Total travel time for LCP 2: ", round(perim(lcp_2)), " m"))

# get average speed
print(paste0("Average speed for LCP 1: ", 
             round(perim(lcp_1)/lcp_1$time, 2), " m/s"))
print(paste0("Average speed for LCP 2: ", 
             round(perim(lcp_2)/lcp_2$time, 2), " m/s"))

```


